Financial Contributions to Presidential Campaigns in California by Miguel Angel Nieto
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(dplyr)
library(gridExtra)
library(scales)
```
# Opening the data set

When I was trying to open the dataset with normal read.csv function I got the 
following error:

**Error in read.table(file = file, header = header, sep = sep, quote = quote,**
**: duplicate 'row.names' are not allowed**
  
I opened the .csv with a normal text editor and counted the number of columns
and values on each row. The number was correct in both cases, but there was
something strange at the end of each row. They had a **comma sign**, so R was
showing an error because there was no one-to-one relantionship.

Using function parameters like **row.names=NULL** fixes the loading process,
but it screws the relationshipt between each column and the corresponding data.
So I finally decided to use normal UNIX tools to fix the data set.
This is what the original data looks like:

**C00575795,"P00003392","Clinton, Hillary Rodham","KLEEMAN, ANNETTE","SANTA MONICA","CA","904021336","N/A","RETIRED",100,04-MAY-16,"","X","* HILLARY VICTORY FUND","SA18","1079219","C5503390","P2016",***

And this is what we need:

**C00575795,"P00003392","Clinton, Hillary Rodham","KLEEMAN, ANNETTE","SANTA MONICA","CA","904021336","N/A","RETIRED",100,04-MAY-16,"","X","* HILLARY VICTORY FUND","SA18","1079219","C5503390","P2016"**

To fix the dataset I did the following:

```$ cat P00000001-CA.csv | sed 's/.$//' > P00000001-CA-fixed.csv```

P00000001-CA-fixed.csv is the file that I am going to use.

```{r echo=FALSE, Load_the_Data}
# Load the Data
ca <- read.csv("P00000001-CA-fixed.csv")

# Leave only the Family Name of the candidate to make graphs easier to read
ca$cand_nm <-(gsub(",.*$", "", ca$cand_nm))

# Add a variable party that will specify to which party a particular
# candidate belongs to.
# First I create the variable just including the names we already have
# (to just fill it)
# and then we use a list to change the values depending on the value each
# object has.

ca$party <- factor(ca$cand_nm)

levels(ca$party) <- list(
    democrat = c("Clinton", "Sanders","O'Malley","Lessig","Webb"),
    republican = c("Cruz", "Rubio", "Bush","Carson","Kasich","Fiorina",
                    "Paul", "Trump","Walker","Christie","Graham",
                    "Huckabee","Perry", "Santorum","Pataki","Jindal",
                    "Gilmore"),
    green = c("Stein"),
    libertarian = c("Johnson")
)
```

# Univariate Plots Section
```{r echo=FALSE}

ggplot(aes(x=cand_nm),data=ca)+
  geom_bar() +
  theme(axis.text.x=element_text(angle = -90, hjust = 0), plot.title = 
          element_text(lineheight=.8, face="bold")) +
  ggtitle("Count of contributions per candidate") +
  ylab("Contributions") +
  xlab("Candidates")

ggplot(aes(x=party),data=ca)+
  geom_bar() +
  theme(axis.text.x=element_text(angle = -90, hjust = 0), plot.title = 
          element_text(lineheight=.8, face="bold")) +
  ggtitle("Count of contributions per party") +
  ylab("Contributions") +
  xlab("Parties")
```

These two graphs show the number of contributions received grouped by different
factors. In first graph we can see that Clinton and Sanders are both the ones
with the largest number of contributions received. Both are democrats, so the
second graph doesn't really suprise us. Democrats are the ones that receive more
contributions in California. Also, taking in account that they won the last two
elections in that State, it shows that the party is really strong there.

```{r echo=FALSE}
# I create different datasets based on their occupation, grouping them in
#retired, studends, not employed and all others

retired <- ca %>%
  filter(contbr_occupation =="RETIRED")

students <- ca %>%
  filter(contbr_occupation =="STUDENT")

not_employed <- ca %>%
  filter(contbr_occupation =="NOT EMPLOYED")

other_people <- ca %>%
  filter(contbr_occupation != "RETIRED" & contbr_occupation != "STUDENT" &
           contbr_occupation != "NOT EMPLOYED")

g_retired <- ggplot(aes(x=cand_nm),data=retired)+
  geom_bar() +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Retired")

g_students <- ggplot(aes(x=cand_nm),data=students)+
  geom_bar() +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Students")

g_not_employed <- ggplot(aes(x=cand_nm),data=not_employed)+
  geom_bar() +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Not employed")

g_other_people <- ggplot(aes(x=cand_nm),data=other_people)+
  geom_bar() +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Others")

grid.arrange(g_retired, g_students, g_not_employed, g_other_people)
```

Here I divided the people in different groups:

- **Retired**
- **Students**
- **Not employed**
- **Others**

We see some trends. For example, seems that Clinton is the favourite of retired
people. While in all other groups Sanders seems to be the winner in the number
of contributions received. Clinton already said that she is against the idea
of raising the retirement age, but I don't think that opinion affect retired
people since it doesn't really affect them anymore. But it is still interesting
to see that retired people are the ones that don't follow what seemst to be the
trend of contributing more times to Sanders.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# List with the top 5 biggest cities in California
biggest_five = c("LOS ANGELES", "SAN DIEGO", "SAN JOSE", "SAN FRANCISCO",
                 "FRESNO")

# Extract the data only from those top 5 biggest cities and store them in a
#new variable

biggest_five_cities <- ca %>%
  filter(contbr_city %in% biggest_five)

ggplot(aes(x=cand_nm),data=biggest_five_cities)+
  geom_bar() +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),plot.title = 
          element_text(lineheight=.8, face="bold")) +
  ggtitle("Number of contributions received per candidate
          in the biggest cities") +
  xlab("Candidates") +
  ylab("Number of contributions")
  
ggplot(aes(x=party),data=biggest_five_cities)+
  geom_bar() +
  theme(plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Number of contributions received per party
          in each of 5 biggest cities") +
  facet_wrap(~contbr_city, scales = "free") +
  ylab("Number of contributions") +
  xlab("Parties")

ggplot(aes(x=cand_nm),data=biggest_five_cities)+
  geom_bar() +
  facet_wrap(~contbr_city, scales = "free") +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Number of contributions received by each candidate
          in the 5 biggest cities") +
  xlab("Candidates") +
  ylab("Number of contributions")

ggplot(aes(x=party),data=biggest_five_cities)+
  geom_bar() +
  facet_wrap(~contbr_city, scales = "free") +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Number of contributions received by each party
          in each of 5 biggest cities") +
  xlab("Parties") +
  ylab("Number of contributions")
```

Investigating the data by City, taking the 5 largest ones, we see again
democrats dominating the numbers being Sanders again the one with more
contributions received. At this moment, looks pretty clear that California
is a Democrat state.

## Summary

- Democrats are clear winnes in the number of contributions received in whole
State and also in each of the 5 biggest cities.
- From all Democrats, Sanders is the winner being Clinton the second one in the
number of contributions received.
- Clinton seems to be the favourite among retired people. All other groups
contribute more times to Sanders.

# Univariate Analysis

### What is the structure of your dataset?

The dataset includes 18 variables with 653397 observations. It includes
contributions made by people to different candidates. People are defined by
their name, the city where they life (and its zip code), ammount of money given,
employer, occupation, receipt date and some other variables used to identify
the contribution itself.

```{r echo=FALSE}
dim(ca)
```

```{r echo=FALSE}
str(ca)
```

The summary of the dataset shows that **Sanders** is the candidate that received
the largest number of contributions, but we still don't know if that means the
largest ammount of money. **PENDERGAST, JAN** is the one with more
contributions, doing it 244 times. Los Angeles is the city with more
contributions and unemployed are the ones that contributed more times.
The median constribution is 27$ and the mean is 126$. There are big ouliners,
for example negative contributions of -10000 (refunds) and positive of 10800
that we would need take in account.

```{r echo=FALSE}
summary(ca)
```

### What is/are the main feature(s) of interest in your dataset?

This dataset is all about money contributions made by people to different
candidates. Usually different group of people, depending on where they live,
occupation (that affects yearly income) have different favourite candidates.
The main features here are the different variables that define people, being
the most interesting ones the **employment status**, **city** where they live
and the **amount of money contributed**. Most of the variables are not really
useful, like transaction ids, forms, file ids... so we are going to concentrate
on those I have already described.

Money helps a lot to win the elections. It provides the candidates with capacity
to use ads, spread propaganda and make their ideas to reach a larger audience.
It could be that the more money they receive, the highest the changes to win.
So, the largest cities and also people with more money could decide who is going
to be the winner. The question is, will the data show that trend?

### What other features in the dataset do you think will help support your
### investigation into your feature(s) of interest?

In this first part I just counted the number of contributions, but that doesn't 
show the full picture. Counting the total ammount of contributions per group 
will start to give us some more information and see what group contributes more
to the winning candidate and which variables seems to have interesting outlines.

### Did you create any new variables from existing variables in the dataset?

I created different dataframes to get data based on the variables I find more
interesting:

- **biggest_five_cities** datafrome, that includes data from top five cities.
That includes Los Angeles, San Diego, San Jose, San Francisco and Fresno
- **not_employed** dataframe for unemployed people
- **students** dataframe for students
- **retired** dataframe for retired people
- **other_people** dataframe for all other employed people

The idea is to be able to use the data based on those particular variables
without having to *subset* it every time I want to use it. That will make the
code easier to write, read and mantain.

I also created the **"Party" variable** to specify to which party the candidate
belongs to.

### Of the features you investigated, were there any unusual distributions?
### Did you perform any operations on the data to tidy, adjust, or change the
### form of the data? If so, why did you do this?

**Unusual distribution and interesting data**

- The two top receiving number of contributions are Hillary Clinton and Bernard
Sanders, both Democrats candidates.

- If we keep investigating the number of contributions we also see that retired
people prefer Hillary Clinton while the other groups go with Bernard Sanders.

- Those without employment, and that usually means without that much money
to spend, are the ones that contributed more times.

- California is a democrat state.

**Operations performed on the data** 

- The *cand_nm* includes the Full Name. I used a regular expression to just
leave the Family Name. That will make graphs easier to read.

# Bivariate Plots Section
```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=cand_nm, y=contb_receipt_amt),data=ca) +
  geom_point(alpha = 1/5,position = "jitter") + 
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Contributions received by candidates") +
  scale_y_continuous(limits = c(0, 15000)) +
  xlab("Candidates") +
  ylab("Contributions received")

ggplot(aes(x=cand_nm, y=contb_receipt_amt),data=ca) +
  geom_bar(stat="identity") + 
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Contributions in total each candidate received") +
  xlab("Candidates") +
  ylab("Total contribution received")

ggplot(aes(x=party, y=contb_receipt_amt),data=ca) +
  geom_bar(stat="identity") +
  theme(plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Contribution received by each party") +
  xlab("Party") +
  ylab("Total contribution received")
```

First graph shows that most of the contributions are bellow 2500$.

Next graphs tells us a totally different story the ones we saw in previous
section. Sanders was the one receiving the largest number of contribution,
but if we count the total amount of money: 
**Clinton is getting nearly twice the money that Sanders receives**.

In the other hand there are no surprises in the last graph. Democrats were the
ones receiving more contributions and we see here they are also the winners in
the total amount of money received.

```{r echo=FALSE, warning=FALSE}
gt_retired <- ggplot(aes(x=cand_nm, y=contb_receipt_amt),data=retired)+
  geom_bar(stat="identity") +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Retired")

gt_students <- ggplot(aes(x=cand_nm, y=contb_receipt_amt),data=students)+
  geom_bar(stat="identity") +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Students")

gt_not_employed <- ggplot(aes(x=cand_nm, y=contb_receipt_amt),
                          data=not_employed)+
  geom_bar(stat="identity") +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Not employed")

gt_other_people <- ggplot(aes(x=cand_nm, y=contb_receipt_amt),
                          data=other_people)+
  geom_bar(stat="identity") +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Others")

grid.arrange(gt_retired, gt_students, gt_not_employed, gt_other_people)
```

Retired people contribute again to Cliton before every other candidate. 
Not employed ones are the only ones that don't follow this trend. Most probably
because Sanders proposed to: 

*expand Social Security, to the tune of $65 more per month on average,*
*financed by raising payroll taxes on wealthy wage earnings. He's also vowed*
*to get the unemployed working again through a $1 trillion infrastructure plan*
*the campaign says will create 13 million "good-paying" jobs.*

[Source](http://fortune.com/2016/02/19/bernie-sanders-unemployed/)

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=cand_nm, y=contb_receipt_amt),data=biggest_five_cities) +
  geom_bar(stat="identity") + 
  theme(plot.title = element_text(lineheight=.8, face="bold"),
        axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),axis.title.y=element_blank()) +
  ggtitle("Total contribution per candidate on the 5 biggest cities") +
  xlab("Candidates") +
  ylab("Total contribution received")

ggplot(aes(x=party, y=contb_receipt_amt),data=biggest_five_cities) +
  geom_bar(stat="identity") + 
  theme(plot.title = element_text(lineheight=.8, face="bold"),
        axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),axis.title.y=element_blank()) +
  ggtitle("Total contribution per party on the 5 biggest cities") +
  xlab("Parties") +
  ylab("Total contribution received")

ggplot(aes(x=cand_nm, y=contb_receipt_amt),data=biggest_five_cities) +
  geom_bar(stat="identity") + 
  facet_wrap(~contbr_city, scales = "free") +
  theme(plot.title = element_text(lineheight=.8, face="bold"),
        axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),axis.title.y=element_blank()) +
  ggtitle("Total contribution per candidate each of the 5 biggest cities")

ggplot(aes(x=party, y=contb_receipt_amt),data=biggest_five_cities) +
  geom_bar(stat="identity") + 
  facet_wrap(~contbr_city, scales = "free") +
  theme(plot.title = element_text(lineheight=.8, face="bold"),
        axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),axis.title.y=element_blank()) +
  ggtitle("Total contribution per party on each of the 5 biggest cities")
```

Here we see even more interesting trendings, that makes our first analysis of
total number of contributions complety useless. In these 5 cities, Sanders was
doubling the number of contributions received. But if we count the money, as we
do in first graph, we see that Clinton gets more than twice money. She is the
clear winner here. Democrats still get much more money (as they got much more
contributions) than other parties.

If we check the graph per city, we also see big differences similar to the one
explained before. In Los Angeles, San Diego, San Francisco and San Jose, the
graphs are nearly the opposite of what we say when we just counted the number
of contributions. Being Clinton again the winner in amount (what really
matters), but not in the number of contributions.

The last thing we see is that Fresno is the one that goes in the opposite
direction from the other 4 biggest cities. Republicans win in the amount of
money contributed and Sanders is the one getting more money.

## Summary

After adding a **y** variable to our graphs (total ammount of money contributed)
we see that the picture changed a lot.

- Democrats are still clear winnes in the number of contributions received in 
whole State, but not in Fresno.
- Sander was the winner in the number of contributions received, but clearly 
Clinton receives the largest total amount of money.
- Retired provided the largest number of contributions to Clinton, and we see
here that also the largest amount of money. Clinton is the winner in all groups
but "not employed" ones. Mostly the opposite of what we say before.

Now, things are even more clear. Democrats are the first ones, and from 
Democrats Clinton is in the first place.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the
### investigation. How did the feature(s) of interest vary with other features
### in the dataset?

If we take in account the candidates, it varied a lot. It is clear that the
number of contributes received doesn't really correlate with the total amount
of money. Sanders clearly gets the largest number of contributions, but
Clinton gets the largest amount of money. Being both Democrats, it is still
the winning party in the state.

### Did you observe any interesting relationships between the other features
### (not the main feature(s) of interest)?

Not employed people prefer Sanders, because of his proposals. Also, Fresno seems
to follow a different pattern from other biggest cities and also the state of
California. Fresno seems to be a Republic city.

### What was the strongest relationship you found?

Not employed people are, with no doubt, supporting Sanders. Seems to be the only
group where Sanders has done a really good job.

# Multivariate Plots Section

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=cand_nm, y=contb_receipt_amt),data=ca) +
  geom_bar(stat="identity",aes(colour=party,fill=party)) + 
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Contributions in total each candidate received") +
  xlab("Candidates") +
  ylab("Total contribution received")

ggplot(aes(x=cand_nm, y=contb_receipt_amt),data=biggest_five_cities) +
  geom_bar(stat="identity",aes(colour=party,fill=party)) + 
  theme(plot.title = element_text(lineheight=.8, face="bold"),
        axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),axis.title.y=element_blank()) +
  ggtitle("Total contribution per candidate on the 5 biggest cities") +
  xlab("Candidates") +
  ylab("Total contribution received")
```

The new layer that adds colour depending on the party who receives the
contribution reinforces the idea that CA is mostly democrat. Republicans'
numbers are pretty low, so it is difficult to get information from them. Let's
remove democrats so we can have an easier to read picture of all others.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=cand_nm, y=contb_receipt_amt),
       data=subset(ca, party != "democrat")) +
  geom_bar(stat="identity",aes(colour=party,fill=party)) + 
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Contributions in total each candidate received") +
  xlab("Candidates") +
  ylab("Total contribution received")
```

Cruz and Rubio are the favourite ones among Republicans.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=cand_nm, y=contb_receipt_amt),data=ca) +
  geom_point(alpha = 1/5,position = "jitter") + 
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Contributions received by candidates") +
  scale_y_continuous(limits = c(0, 15000)) +
  xlab("Candidates") +
  ylab("Contributions received") +
  geom_hline(yintercept=mean(ca$contb_receipt_amt), color="red",
             linetype="dashed")
```

The mean line added shows that there are many ouliners here making it a bit
useless without some adjustments first. 

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=cand_nm, y=contb_receipt_amt),data=ca) +
  geom_point(alpha = 1/100, position = "jitter") + 
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        plot.title = element_text(lineheight=.8, face="bold")) +
  ggtitle("Contributions received by candidates") +
  scale_y_continuous(limits = c(0, 1000)) +
  xlab("Candidates") +
  ylab("Contributions received") +
  geom_hline(yintercept=mean(ca$contb_receipt_amt), color="red",
             linetype="dashed")
```

Limiting the y from 0 to 1000 and adding an alpha of 1/100 we can get a better
picture. Clinton and Sanders have so high number of contributions that they are
really above the mean. Both candidates are per-se ouliners. We see that people
tend to contribute with 100$, 250$, 500$, 750$ and 1000$. Those are round
numbers, so they appear often.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the
###investigation. Were there features that strengthened each other in terms of
###looking at your feature(s) of interest?

There is not much new data, since we have already explored most of the useful
variables of the data set. Adding parties as a third variable in our graphs
helps us to see the big difference between Democrats and Republicans in
California. Something we already found before but gets even more clear now.

### Were there any interesting or surprising interactions between features?

Not really. The data we have seen was the expected one after previous analysis.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, warning=FALSE, Plot_One}
ggplot(aes(x=cand_nm, y=contb_receipt_amt),data=ca) +
  geom_bar(stat="identity",aes(colour=party,fill=party)) + 
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        plot.title = element_text(lineheight=.8, face="bold")) +
  geom_point(aes(y=contb_receipt_amt)) +
  ggtitle("Contributions in total each candidate received") +
  scale_y_continuous(name="Total contribution", labels = dollar) +
  xlab("Candidates")
```

### Description One

The data we are analysing includes the amount of money each candidate has received
from individual contributors in California. Therefore, the first graph shown here
is created to get a first overview of general contributions in California and try
to get an overall idea. It shows the total amount of dolars received by each
candidate.

In the graph we can see the total amount of dolars received by each candidate. 
Colors represent the party to which the candidate belongs to. Clinton and 
Sanders, both democrat, are the ones getting more financial help. The difference
between the two candidates is also pretty large with Clinton receiving almost
twice more money than Sanders.

**Contributions received by each candidate**
```{r echo=FALSE}
by(ca$contb_receipt_amt, ca$cand_nm, sum)
```

This graph also shows that democrat party is dominant in California with both
candidates acting as ouliners in the graph. The rest of candidates' numbers
really low in comparison.

**Number of contributions received by each party**
```{r echo=FALSE}
table(ca$party)
```

Even if the graph looks pretty much straightforwad, we cannot just extrapolate
the data to every single combination and group of people so in
next graph I am going to show some extreme cases that actually show a
complety different picture.

### Plot Two
```{r echo=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(x=cand_nm, y=contb_receipt_amt),
       data=subset(not_employed, cand_nm == "Clinton" | cand_nm == "Sanders"))+
  geom_bar(stat="identity",aes(color=cand_nm, fill=cand_nm)) +
  theme(axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),axis.ticks=element_blank(),
        plot.title = element_text(lineheight=.8, face="bold"),
        legend.position="none") +
  ggtitle("Not employed") +
  scale_y_continuous(name="Total contribution", labels = dollar)
```

### Description Two

The data set includes information about the employment of each contributor. This
is a very important piece of information, because people tend to vote based on
their personal situation. So, it is pretty usual to see different groups 
supporting different candidates depending on the wealth or job.

I have divided the contributors in different groups:

- Unemployed
- Students
- Retired
- All others

The idea is to find some group that doesn't follow the general picture saw
in previous graph. So, in this second picture we can see that unemployed people
are the only group that prefer Sanders over Clinton. The difference is also
very large. 

**Contributions received by Sanders in total from everybody:**
```{r echo=FALSE}
by(subset(ca, cand_nm=="Sanders")$contb_receipt_amt, 
subset(ca, cand_nm=="Sanders")$cand_nm, sum)
```

**Contributions received by Sanders in total from unemployed:**
```{r echo=FALSE}
by(subset(not_employed, cand_nm=="Sanders")$contb_receipt_amt, 
subset(not_employed, cand_nm=="Sanders")$cand_nm, sum)
```

Sanders receives 26% of contributions from this group. More information about
why this could happen in *reflection* section.

### Plot Three
```{r echo=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x=party, y=contb_receipt_amt),
       data=subset(biggest_five_cities,contbr_city=="FRESNO")) +
  geom_bar(stat="identity",aes(color=party,fill=party)) + 
  theme(plot.title = element_text(lineheight=.8, face="bold"),
        axis.text.x=element_text(angle = -90, hjust = 0),
        axis.title.x=element_blank(),axis.title.y=element_blank()) +
  ggtitle("Total contribution per party in Fresno") +
  xlab("Parties") +
  scale_y_continuous(name="Total contribution", labels = dollar)
```

### Description Three

Usually biggest cities have the biggest impact, just because of the population.
So I have analyzed the data from the top 5 biggest cities in California and in 
the same way I did in *graph two* I tried to find something that goes against 
the ideas the *first graph* showed us.

In fact, we can see in this graph that Fresno is not democrat. Republican
candidates get most of the contributions, being a totally different picture
from the overall data of California. Actually, Fresno's state and federal
representation is mostly republican as we can see in the wikipedia links
presented in next section.

------

# Reflection

I don't live in the United States, so I started this project without background
knowledge of their democratic system or preconceived ideas. While analysing the
data I learn some very interesting information, summarized in the last three 
graphs from previous section. I used google to find news articles and wikipedia
information to find that what I was discovering just checking the graphs was
the reality.

- Clinton is the favourite one among democrats.
https://en.wikipedia.org/wiki/Nationwide_opinion_polling_for_the_Democratic_Party_2016_presidential_primaries

- Sanders is the favourite one among non employeed people.
https://heatst.com/politics/bernie-sanders-has-the-most-unemployed-donors/

- CA is a Democrat state.
https://en.wikipedia.org/wiki/California

- Fresno goes in opposite direction from other big cities from the state. 
They are republicans.
https://en.wikipedia.org/wiki/Fresno,_California

It is really amazing to see how you can discover real life facts and realities
just by graphing numbers. Checking the selected three graphs also tell us
something really important. The full picture, the graph of whole California,
can't be used as a perfect representation of each single city and group of 
people individually. There will be some that doesn't follow the general rule
and each because of different reasons.

I have encountered some problems while working with the data set. As I
mentioned, most of the data is mostly useless and cannot be correlated with
each other variable. Zip codes, candidate Ids, memo_cd, memo_text, file number
and so on. So, I had to investigate only a small subset of those variables. To
get some more information, I divided employment in groups and added the party
of each candidate, so I could get some more data. But still, there were not many
combinations. Actually taking in account that the topic was contributions made
to politics, there is no much more information needed apart from money and some
other metadata from the donors.

I was able to extract the data I was interested in and I learnt a lot about 
California, the different candidates and each political party.

For the future it could be even better if there was data about really big 
donations from enterprises. That would help us to see the usual relations 
between gas/oil/energy/technology companies and different candidates based on 
their ideas and future plans. Those big companies are the ones that really rule 
the world and their contributions will change it in a more drastic way 
that individuals and their donations.
